A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-27 paper1_tKdV/tKdV_theory.tex
NM

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. 
Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). Downstream of the ADC, PDFs of  surface displacement exhibit strong positive skewness, which is accompanied by an elevated level of extreme events. 
Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media.
\end{abstract}

@@

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important characterizing parameters measured from the experiments
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength scale $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$
the normalized wave depth ratio with incoming flow depth $d=H_{0}$
to the outgoing flow depth $d<H_{0}$. The interpretations and reference
values of these model parameters are based on the experimental setup
\cite{bolles2018anomalous}. By comparing the characteristic physical
scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked directly with the measured non-dimensional quantities by
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ represents the factor to normalize the total
energy in the state variable $u_{\Lambda}$ to one.

Consider the spatial discretization $J=2\Lambda+1$ so that the smallest
resolved scale is comparable with the characteristic wavelength
\[
\Delta x=\frac{2\pi M\lambda_{c}}{J}\lesssim\lambda_{c}\:\Rightarrow\:M=\frac{J}{2\pi}\sim5,\;J=32.
\]
Therefore in the practical numerical simulations, we pick $M=5$ and
$\gamma$ varies in the range $\left[0.5,1\right]$. Using the reference
experimental measurements \cite{bolles2018anomalous}, $\epsilon\in\left[0.0024,0.024\right],\delta\sim0.22$,
and $D_{0}$ changes from 1 to 0.24 before and after the depth change.
The reference values for the model scales can be estimated in the
range $L_{0}\in\left[2,6\right]$ and $E_{0}\in\left[50,200\right]$.
These are the values we will test in the direct numerical simulations.
See details about the derivation from scale analysis in \textcolor{blue}{\emph{SI Appendix, A}}.

>>>>

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. 
Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). Downstream of the ADC, PDFs of  surface displacement exhibit strong positive skewness, accompanied by an elevated level of extreme events. 
Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media.
\end{abstract}

% ORIGINAL SENTENCE IN ABSTRACT
%Recent controlled laboratory experiments in weakly turbulent shallow water with abrupt depth change (ADC) exhibit a remarkable transition from nearly Gaussian statistics in incoming wave trains before the ADC to outgoing waves trains after the ADC with extreme anomalous statistics with large positive skewness of the surface height.

@@

\cite{bolles2018anomalous}

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important parameters that characterize the experiments
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to (upstream) water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$ 
the depth ratio with upstream value $d=H_{0}$
and downstream value $d<H_{0}$. 
By comparing the characteristic physical scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked directly to the above experimental parameters via
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ normalizes the total
energy of the state variable $u_{\Lambda}$ to one.
\todo[inline]{What is the variable $U$? Have we defined it?}

Consider a spatial discretization with $J=2\Lambda+1$ grid points, so that the smallest
resolved scale is comparable to the characteristic wavelength
\[
\Delta x=\frac{2\pi M\lambda_{c}}{J}\lesssim\lambda_{c}\:\Rightarrow\:M=\frac{J}{2\pi}\sim5,\;J=32.
\]
In the practical numerical simulations, we select $M=5$ and
$\gamma$ varies in the range $\left[0.5,1\right]$. Using the reference
experimental measurements \cite{bolles2018anomalous}, $\epsilon\in\left[0.0024,0.024\right],\delta\sim0.22$,
and $D_{0}$ changes from 1 to 0.24 before and after the depth change.
The reference values for the model scales can be estimated in the
range $L_{0}\in\left[2,6\right]$ and $E_{0}\in\left[50,200\right]$.
These are the values we will test in the direct numerical simulations.
See details about the derivation from scale analysis in \textcolor{blue}{\emph{SI Appendix, A}}.
